---
title: "STA302 Group 85 Project Proposal"
authors: 'Kiarash Kianidehkordi,Joyce Lin,Min Ji koo '
output:
  pdf_document: default
  html_notebook: default
---

#Assumptions of linear regression
Y is related to x by the linear regression model
Yi=B0+B1xi+ei
the errors e1...ei are independent of each other
the errors e1,ei have common variance of sigma^2
the errors are normally disturbuted with a mean of 0 and variance of sigma^2



#Installing relevant packages and libraries
```{r}
install.packages("tidyverse")
install.packages("car")
install.packages("readr")
install.packages("dplyr")
install.packages("here", dependencies = TRUE)
install.packages("stargazer")

library(readr)
library(dbplyr)
library(MASS)
library(car)
library(here)
library(stargazer)

```

```{r}
totalData <-read_csv(here("data", "hour.csv"))

totalData$is_winter <- ifelse(totalData$season == 1, 1, 0)
totalData$season <- NULL

head(totalData)

str(totalData)

summary(totalData)
#extract the hour data for peak hours 
hourData <- totalData[totalData$hr %in% c(8, 17, 18) & totalData$yr == 1, ]


# Scatter plot of bike rental count vs season
plot(hourData$is_winter, hourData$cnt,
     main = "Bike Rental Count vs Season",
     xlab = "Season",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)

boxplot(hourData$cnt ~ hourData$is_winter,
        main = "Bike Rental Count by Season",
        xlab = "Season",
        ylab = "Bike Rental Count",
        col = "lightgreen",
        names = c("Not Winter", "Winter"))
# Scatter plot of bike rental count vs wind speed
plot(hourData$windspeed, hourData$cnt,
     main = "Bike Rental Count vs Wind Speed",
     xlab = "Wind Speed",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)
# Scatter plot of bike rental count vs temperature
plot(hourData$temp, hourData$cnt,
     main = "Bike Rental Count vs Temperature",
     xlab = "Temperature",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)

# Scatter plot of bike rental count vs humidity
plot(hourData$hum, hourData$cnt,
     main = "Bike Rental Count vs Humidity",
     xlab = "Humidity",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)
# Scatter plot of bike rental count vs hour of the day
plot(hourData$hr, hourData$cnt,
     main = "Bike Rental Count vs Hour of the Day",
     xlab = "Hour of the Day",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)
boxplot(hourData$cnt ~ hourData$workingday,
        main = "Bike Rental Count by Working Day",
        xlab = "Working Day",
        ylab = "Bike Rental Count",
        col = "lightcoral",
        names = c("Non-Working Day", "Working Day"))

```
```{r}
# Fit the linear model with Winter and Working Day as binary indicators
initialFitted <-lm (cnt ~ temp +  windspeed + hum + is_winter + I(workingday == 1), data = hourData)
# Display the summary of the model
summary(initialFitted)

# Visualize diagnostics
par(mfrow = c(1, 2))  # Arrange plots side by side


# 2. Normal Q-Q plot
plot(initialFitted, which = 2)



```

```{r}
# Set the desired dimensions for the image (in pixels)
width <- 600   # Width in pixels
height <- 600  # Height in pixels (adjust based on aspect ratio)

# Open a PNG device to save the plot
png("IMAGES/residuals_and_outliers_custom.png", width = width, height = height)

# Set up a 2x2 plotting layout
par(mfrow=c(2, 2))

# Generate the plots (this will create 4 diagnostic plots)
plot(initialFitted, which = c(1, 2, 3, 4))

# Turn off the device to save the image
dev.off()

```

```{r}
cooks.distance(initialFitted)[c(230, 459, 543)]
```
```{r}
hourData[c(230, 459, 543), ]
```
```{r}
rstandard(initialFitted)[c(230, 459, 543)]

hatvalues(initialFitted)[c(230, 459, 543)]

dfbetas(initialFitted)[c(230, 459, 543), ]
dffits(initialFitted)[c(230, 459, 543)]
```


#Analysis of regression based on the proposed procedure in part 2 
The procedure followed in this analysis is available at [Sta302 Project Part 2](https://github.com/kiarashkianid/Washington-Bike-Sharing-Regression-analysis/blob/main/STA302%20project%20part%202.drawio%20(1).png).

#Step 1
we will use the initialFitted model as our starting point. To check the assumption of linear regression our first task is to draw residual vs fitted plot and residual vs predicors scatter plot 
```{r}
# Select relevant predictors for the matrix plot
predictors <- hourData[, c("temp", "hum", "windspeed","is_winter","workingday")]

png("IMAGES/matrix_plot_predictors.png", width = 800, height = 600)

# Create a matrix plot
pairs(predictors,
      main = "Matrix Plot of Predictors",
      col = "lightblue",
      pch = 19)

dev.off()
```
#we also draw the corrolation matrix to see the relations among the predictors and response.
```{r}
# Select relevant numerical predictors for correlation analysis
numericalData <- hourData[, c("temp", "hum", "windspeed","workingday","is_winter","cnt")]

# Calculate the correlation matrix
corMatrix <- cor(numericalData)

# Print the correlation matrix
print(corMatrix)

# Install and load the corrplot package
install.packages("corrplot")
library(corrplot)

# Create a correlation plot
corrplot(corMatrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, # Rotate labels for better readability
         title = "Correlation Matrix of Variables", 
         mar = c(0,0,1,0))  # Adjust margins for better display

```
as we see count has highest correlation with temperature, humidity and hour of the day and least correlation with wind speed and whether the day is working day
#Step 2: the residual vs fitted plot and the residual vs predictors scatter plots
```{R}
# Set the graphics device to save as PNG
png("IMAGES/residuals_plots.png", width = 800, height = 800)

# Set up a 3x2 plotting layout
par(mfrow = c(3, 2))

# 1. Residuals vs Fitted plot
plot(initialFitted, which = 1)

# Calculate residuals
residuals <- initialFitted$residuals

# Set up the plotting area to display multiple plots

# 1. Residuals vs Temperature
plot (hourData$temp, residuals,
     main = "Residuals vs Temperature",
     xlab = "Temperature",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

# 2. Residuals vs Windspeed
plot( hourData$windspeed, residuals,
     main = "Residuals vs Windspeed",
     xlab = "Windspeed",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

# 3. Residuals vs Humidity
plot(hourData$hum, residuals,
     main = "Residuals vs Humidity",
     xlab = "Humidity",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)


# 5. Box plot for Residuals vs Winter (1 = Yes, 0 = No)
boxplot(residuals ~ (hourData$is_winter),
        main = "Residuals vs Winter",
        xlab = "Winter (1 = Yes, 0 = No)",
        ylab = "Residuals",
        col = c("skyblue", "lightcoral"),
        names = c("No (0)", "Yes (1)"),
        border = "darkblue",
        notch = TRUE) # Adds notches to visualize median differences

# 6. Box plot for Residuals vs Working Day (1 = Yes, 0 = No)
boxplot(residuals ~ hourData$workingday,
        main = "Residuals vs Working Day",
        xlab = "Working Day (1 = Yes, 0 = No)",
        ylab = "Residuals",
        col = c("lightgreen", "orange"),
        names = c("No (0)", "Yes (1)"),
        border = "darkgreen",
        notch = TRUE) # Adds notches to visualize median differences

dev.off()
# Reset the plotting layout to a single plot
par(mfrow = c(1, 1))
```
- Observing residual vs fitted plot of the model , we see that the residuals show a reverse funnel pattern for fitted values within (-100-250+-20) and a scattered pattern from 250+.
More so we make the following observations on the single predictor vs residual plots : 


-  residual vs Temperature : most residuals fall within the 0.2-0.8 temperatures and the plot has a bell shape
-  residual vs Wind speed : we see a funnel pattern with most residuals being in the lower wind speeds.
-  residual vs Humidity :  residuals are scattered with no observable patterns
For residuals vs dummy variables of isWinter and isWorking day :
-  residual vs Winter : we observe both winter and not winter residuals are distributed around zero with IQR for not-winter residuals being (400:-400)) while IQR for winter residuals being (-200: 200). both
-  residual vs working day : we observe both isworkingday and isnotworking day residuals are distributed around zero with IQR range [400,-400] and [300,-300] respectively , with isworking day having multiple outliers

#Step 3 
Since a pattern seems to be present in residual vs predictors and residual vs fitted , we need to add new terms or transform predictor variables or response variables. We will start by checking for multi-col linearity  and non linearity of predictors.


```{R}
#find the VIF to check if multi-col linearity is present
vif_values <- vif(initialFitted)

# Print VIF values
print(vif_values)
```
#Linearity assumption checking
Since VIF values are less than 5 , the model predictors seem to be acceptable(weak multi collinearity )
Now need to address the assumptions of linear regression before we can conduct variable selection. we will start by linearity assumption. 
-we check residuals against each predictor to identify specific non-linear relationships. Residual vs Temperature seems not to be problematic as it show a pattern that indicates the current model is  explaining the relation among predictors. 
# - we will add I(hum^2) since the residual vs humidity plot seems to follow a non linear pattern.
   - uh temperature seems more non-linear than humidity?

resulting model after the transformation show some improvement with regard to linearity assumption but we still need to address both heteroscedasticity and normality assumptions.
```{R}
transformedFitted <- lm (cnt ~ temp + windspeed  + (I(hum^2)) + is_winter + I(workingday == 1), data = hourData)




# 3. Residuals vs Humidity^2
plot( (I(hourData$hum^2)), transformedFitted$residuals,
     main = "Residuals vs Humidity^2",
     xlab = "Humidity^2",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)
# 3. Residuals vs Temp
plot( (hourData$temp), transformedFitted$residuals,
     main = "Residuals vs Temp",
     xlab = "Humidity",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

```

# Constant variance assumption checking
residuals vs. predictor plots indicate that residuals are centered around zero with no discernible pattern  suggesting  that the linearity assumption is met. 

Now we need to address the heteroscedasticity present. We first try to transform response only (using log) the decision is due to the fact that log transformation is used for cases where variance of residual increase with the number of fitted value. however after applying the log transformation we see that the residuals go from a reverse funnel to a funnel pattern (decreasing variance). 

```{R}

plot(transformedFitted, which = 1)


transformedFitted2 <- lm (log(cnt) ~ temp+ windspeed  + I(hum^2) + is_winter + I(workingday == 1), data = hourData)
plot(transformedFitted2, which = 1)


```

this suggest that the log transformation is over correcting the heteroscedasticity. we will try sq-rt transformation on the response. which slightly improve our model. since transforming only the response does not address the non constant variance present in the residuals we will try to transform our predictors. 

```{R}



transformedFitted3 <- lm (sqrt(cnt) ~ temp+ windspeed  + I(hum^2) + is_winter + I(workingday == 1), data = hourData)
plot(transformedFitted3, which = 1)


```



- the plot of residuals vs temperature seems to satisfy the homoscedasticity so we don't transform temperature and we keep the temp as it is 

- we transform wind speed due to a funnel pattern present, we use a log transformation(wind speed+1), +1 is for not taking log(0) which slightly improve the funnel shape.

- we also transform hum^2 by taking a log transformation
```{R}


# 3. Residuals vs Humidity^2
plot( (I(hourData$hum^2)), transformedFitted$residuals,
     main = "Residuals vs Humidity^2",
     xlab = "Humidity^2",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)
# 3. Residuals vs windspeed
plot( (hourData$windspeed) , transformedFitted3$residuals,
     main = "Residuals vs windspeed",
     xlab = "windspeed",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)
transformedFitted4 <- lm (sqrt(cnt) ~ temp+ log(windspeed+1)  + log(I(hum^2))  + is_winter + I(workingday == 1), data = hourData)
plot(transformedFitted4, which = 1)
# 3. Residuals vs log(Humidity^2)
plot( log(I(hourData$hum^2)), transformedFitted$residuals,
     main = "Residuals vs log(Humidity^2)",
     xlab = "log(Humidity^2)",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)
# 3. Residuals vs windspeed
plot( (log(hourData$windspeed+1)) , transformedFitted4$residuals,
     main = "Residuals vs log (windspeed+1)",
     xlab = "log (windspeed+1)",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)


```



at this point the model we have is : 

sqrt (count) = Temp + log(wind speed+1) + log(Humidity^2) + is Winter + is Working day
 
#Normality assumption checking

looking at QQ plots we see that residuals deviate from normality hence we implement box cox to transform 

```{R}

# Perform Box-Cox transformation
boxcox_result <- boxcox(transformedFitted4, lambda = seq(-4, 4, by = 0.1))


```
the first iteration suggest that log-likelihood is maximized when lambda =2 , that is we need to transform the response by a power of 2 , undo sqrt transformation to make residuals normal.

Now that we have addressed the assumptions of linear regression for the initial model we procede with F test for overall model significance 
```{R}

transformedFitted5 <- lm (cnt ~ temp+ log(windspeed+1)  + log(hum^2) + is_winter + I(workingday == 1), data = hourData)

vif(transformedFitted5)

summary(transformedFitted5)
```
since the p-value for the F test is less than the 0.05 significance threshold we reject the null hypothesis in favor of the alternative. That is to conclude that the model as a whole is significant and at least one of the predictors in the model is significantly related to bike rentals counts.


Now we proceed with variable selection which drops the wind speed as a predictor.

```{R}
# Perform backward selection based on BIC
backwardBICModel <- step(transformedFitted5, direction = "backward", k = log(nrow(hourData)))

# Display the summary of the selected model
summary(backwardBICModel)

# Plot the residuals vs fitted values for the final model
plot(backwardBICModel, which = 1)  # Residuals vs Fitted

# Plot normal Q-Q plot
plot(backwardBICModel, which = 2)  # Normal Q-Q plot

```

```{r}


```




