---
title: "Should Washington DC's Bikeshare System Close During the Winter Months?"
author: "Kiarash Kianidehkordi, Min Ji Koo, Joyce Lin"
date: "`r Sys.Date()`"
output: html_document
---

# Contributions

- **Kiarash Kianidehkordi**:  
- **Min Ji Koo**:  
- **Joyce Lin**:  

# Introduction

In the United States, motor transportation accounts for 20% of greenhouse gas emissions (Agency, 2005; Gotschi & Mills, 2008). Bike-sharing systems represent a new generation of traditional bike rentals, where the processes of membership, rental, and return have been automated. Integrating bike-sharing programs into urban transportation networks is considered as a sustainable approach to reducing air pollution (CO2 emissions) and conserving natural resources (Bajracharya, Mulya, Purbasari, & Hwang, 2018; Cai et al., 2019; Lumsdon & Tolley, 2001; Zhang, Zhang, Duan, & Bryde, 2015).

This integration contributes to a greener environment, promotes economic transportation by alleviating traffic congestion and noise pollution, and fosters healthier communities by addressing issues like diabetes and obesity(Shaheen et al. 2013). Environmental and logistical factors are common metrics that researchers look at to determine the optimal allocation of bikes.

Numerous studies indicate a strong correlation between climate conditions and bike-sharing demand. Seasonal fluctuations in bike-sharing usage are often attributed to weather factors (Fournier, Christofa, and Knodler, 2017; Wadud, 2014). For instance, many bike-sharing programs, such as Great Rides, Capital Bikeshare, and Boulder B-cycle, cease operations in winter due to lower trip generation (Godavarthy & Taleqani, 2017). Given this context, the paper aims to investigate if **it is statistically justifiable for Washington DC’s bike-sharing system to reassess its operational strategy during the winter months due to a significant decline in bike sharing usage frequency?**

# Data Description

The bike sharing dataset, extracted from the UCI Machine Learning Repository (Fanaee-T, H. 2013), contains two years' worth of data from the Capital Bikeshare (CBS) system. In the CBS system, when a rental occurs, the software collects basic trip information such as duration, start date, end date, start station, end station, bike number, and member type (Fanaee-T, H., Gama, J 2013).

To avoid trend issues, the authors of the original paper (Fanaee-T, H., Gama, J, 2013) selected data corresponding only to the years 2011 and 2012, consisting of 3,807,587 records. After collecting the raw trip data, the authors aggregated it at two levels: hourly and daily. The hourly time series contains 17,379 hours, and the daily time series contains 731 days (Fanaee-T, H., Gama, J, 2013). Both time series were split into two sets: 2011 (train) and 2012 (test).

Weather temperature, apparent temperature, wind speed, wind gust, humidity, pressure, dew point, and visibility for each hour from January 1, 2011, to December 31, 2012, for Washington, D.C., USA, were extracted from (Freemeteo 2013) and mapped to the corresponding hours in the bike rental time series.

For hours with missing weather reports, the closest available report was used. Additionally, the official holidays of Washington, D.C., were extracted and mapped to the corresponding dates from (Department of Human Resources 2013). Holidays were combined with weekends, and each day was classified as either a working day or a non-working day. Based on the weather data, each hour was assigned one of four weather grades: good, cloudy, bad, or very bad (Fanaee-T, H., Gama, J,2013).

In the hourly scale, each record includes variables such as hour, month, working day status, season, weather grade, temperature, felt temperature, humidity, and wind speed, along with the hourly aggregated count of rented bikes as the target value. In the daily scale, each record contains variables such as month, working day status, season, daily average weather grade, daily average temperature, daily average felt temperature, daily average humidity, and daily average wind speed.

# Preliminary Results

**Data Source**: Bikeshare data from Washington DC, covering 2011-2012.

- **Variables used in this analysis**:
  - `season`: season (1: winter, 2: spring, 3: summer, 4: fall)
    - `is_winter`: variable created from season (1: winter, 0 otherwise)
  - `workingday`: weekday that is not a holiday (1, 0 otherwise)
  - `temp`: normalized temperature (°C), calculated by (t - t_min) / (t_max - t_min)
  - `hum`: normalized humidity. The values are divided to 100 (max)
  - `windspeed`: Normalized wind speed. The values are divided to 67 (max)
  - `cnt`: count of total rental bikes including both casual and registered

- **Other variables**
  - `dteday`: date variable (YYYY-MM-DD)
  - `hr`: hour of the day (0-23)
  - `mnth`: month variable (1-12)
  - `yr`: year (0: 2011, 1: 2012)
  - `weathersit`: discrete variable denoting the weather conditions during a given hour
  - `atemp`: Normalized feeling temperature in Celsius

I will note that while this dataset has time series elements, the research question focuses on the effect of *winter specifically* on Bikeshare usage.  Thus, I believe a linear regression is still appropriate with this research question, as the analysis will be focusing on the effect of winter = 1 versus winter = 0.

This analysis will take data from 2011 and for the hours 5 - 6PM and 8AM.  The year limiter is because only 2011 data is intended to be training data, while  5 - 6PM and 8AM are shown to be the top 3 busiest hours for Bikeshare usage.  Focusing on peak hours in the morning and the afternoon captures the most critical hours for bike share demand and removes temporal continuity that would be present if all hours of the day were considered instead, which would have made this analysis invalid for a linear regression.

Of course, these choices hinge on the assumption that the variation of Bikeshare usage during the top 3 hours, holding other variables like temp and workingday constant, is independent.

```{r}
# Load libraries and data
library(ggplot2)
library(readr)
library(tidyverse)
data <- read.csv("hour.csv")
data$dteday <- as.Date(data$dteday)
summary(data)
```
```{r}
# Group the data by hour and calculate the average 'cnt' (bike usage)
avg_usage_by_hour <- aggregate(cnt ~ hr, data = data, FUN = mean)

# Sort by average usage in descending order and select the top 3 hours with the highest average usage
avg_usage_by_hour <- avg_usage_by_hour[order(-avg_usage_by_hour$cnt),]
top_3_hours <- head(avg_usage_by_hour, 3)
print(top_3_hours)
```


constant variance and linearity assumptions are violated.

Create QQplot to see distribution of errors (analyze normality)






