---
title: "STA302 Group 85 Project Proposal"
authors: "Kiarash Kianidehkordi,Joyce Lin,Min Ji koo "
output: html_notebook
---

#Assumptions of linear regression
Y is related to x by the linear regression model
Yi=B0+B1xi+ei
the errors e1...ei are independent of each other
the errors e1,ei have common variance of sigma^2
the errors are normally disturbuted with a mean of 0 and variance of sigma^2



#Installing relevant packages and libraries
```{r}
install.packages("tidyverse")
install.packages("readr")
install.packages("dplyr")

library(readr)
library(dbplyr)
library(MASS)
```

```{r}
totalData <-read_csv("C:\\Users\\Kiarash\\OneDrive\\Desktop\\University\\statistics\\STA302\\bike+sharing+dataset\\hour.csv")
head(totalData)

str(totalData)

summary(totalData)
hourData<-totalData[totalData$yr==1,]
#plot(hourData[, c(1,3,11,17)], col="cadetblue")


# Scatter plot of bike rental count vs season
plot(hourData$season, hourData$cnt,
     main = "Bike Rental Count vs Season",
     xlab = "Season",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)

boxplot(hourData$cnt ~ hourData$season,
        main = "Bike Rental Count by Season",
        xlab = "Season",
        ylab = "Bike Rental Count",
        col = "lightgreen",
        names = c("Winter", "Spring", "Summer", "Fall"))
# Scatter plot of bike rental count vs wind speed
plot(hourData$windspeed, hourData$cnt,
     main = "Bike Rental Count vs Wind Speed",
     xlab = "Wind Speed",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)
# Scatter plot of bike rental count vs temperature
plot(hourData$temp, hourData$cnt,
     main = "Bike Rental Count vs Temperature",
     xlab = "Temperature",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)

# Scatter plot of bike rental count vs humidity
plot(hourData$hum, hourData$cnt,
     main = "Bike Rental Count vs Humidity",
     xlab = "Humidity",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)
# Scatter plot of bike rental count vs hour of the day
plot(hourData$hr, hourData$cnt,
     main = "Bike Rental Count vs Hour of the Day",
     xlab = "Hour of the Day",
     ylab = "Bike Rental Count",
     col = "lightblue",
     pch = 19)
boxplot(hourData$cnt ~ hourData$workingday,
        main = "Bike Rental Count by Working Day",
        xlab = "Working Day",
        ylab = "Bike Rental Count",
        col = "lightcoral",
        names = c("Non-Working Day", "Working Day"))

```
```{r}
# Fit the linear model with Winter and Working Day as binary indicators
initialFitted<-lm( cnt~ temp + windspeed + hum  + hr + I(season == 1) + I(workingday == 1), data = hourData)

# Display the summary of the model
summary(initialFitted)

# Visualize diagnostics
par(mfrow = c(1, 2))  # Arrange plots side by side


# 2. Normal Q-Q plot
plot(initialFitted, which = 2)



```


#Analysis of regression based on the proposed procedure in part 2 
The procedure followed in this analysis is available at [Sta302 Project Part 2](https://github.com/kiarashkianid/Washington-Bike-Sharing-Regression-analysis/blob/main/STA302%20project%20part%202.drawio%20(1).png).

#Step 1
we will use the initialFitted model as our starting point. To check the assumption of linear regression our first task is to draw residual vs fitted plot and residual vs predicors scatter plot 
```{r}
# Select relevant predictors for the matrix plot
predictors <- hourData[, c("temp", "hum", "windspeed","season","workingday")]

# Create a matrix plot
pairs(predictors,
      main = "Matrix Plot of Predictors",
      col = "lightblue",
      pch = 19)

```
#we also draw the corrolation matrix to see the relations among the predictors and response.
```{r}
# Select relevant numerical predictors for correlation analysis
numericalData <- hourData[, c("temp", "hum", "windspeed", "hr","workingday","season","cnt")]

# Calculate the correlation matrix
corMatrix <- cor(numericalData)

# Print the correlation matrix
print(corMatrix)

# Install and load the corrplot package
install.packages("corrplot")
library(corrplot)

# Create a correlation plot
corrplot(corMatrix, method = "circle", type = "upper", 
         tl.col = "black", tl.srt = 45, # Rotate labels for better readability
         title = "Correlation Matrix of Variables", 
         mar = c(0,0,1,0))  # Adjust margins for better display

```
as we see count has highest correlation with temperature, humidity and hour of the day and least correlation with wind speed and whether the day is working day
#Step 2: the residual vs fitted plot and the residual vs predictors scatter plots
```{R}
# 1. Residuals vs Fitted plot
plot(initialFitted, which = 1)

# Calculate residuals
residuals <- initialFitted$residuals

# Set up the plotting area to display multiple plots

# 1. Residuals vs Temperature
plot(hourData$temp, residuals,
     main = "Residuals vs Temperature",
     xlab = "Temperature",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

# 2. Residuals vs Windspeed
plot(hourData$windspeed, residuals,
     main = "Residuals vs Windspeed",
     xlab = "Windspeed",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

# 3. Residuals vs Humidity
plot(hourData$hum, residuals,
     main = "Residuals vs Humidity",
     xlab = "Humidity",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

# 4. Residuals vs Hour of the Day
plot(hourData$hr, residuals,
     main = "Residuals vs Hour of the Day",
     xlab = "Hour of the Day",
     ylab = "Residuals",
     col = "lightblue",
     pch = 19)

# 5. Box plot for Residuals vs Winter (1 = Yes, 0 = No)
boxplot(residuals ~ (hourData$season == 1),
        main = "Residuals vs Winter",
        xlab = "Winter (1 = Yes, 0 = No)",
        ylab = "Residuals",
        col = c("skyblue", "lightcoral"),
        names = c("No (0)", "Yes (1)"),
        border = "darkblue",
        notch = TRUE) # Adds notches to visualize median differences

# 6. Box plot for Residuals vs Working Day (1 = Yes, 0 = No)
boxplot(residuals ~ hourData$workingday,
        main = "Residuals vs Working Day",
        xlab = "Working Day (1 = Yes, 0 = No)",
        ylab = "Residuals",
        col = c("lightgreen", "orange"),
        names = c("No (0)", "Yes (1)"),
        border = "darkgreen",
        notch = TRUE) # Adds notches to visualize median differences

# Reset the plotting layout to a single plot
par(mfrow = c(1, 1))

```
- Observing residual vs fitted plot of the model , we see that the residuals show a reverse funnel pattern for fitted values within (-100-250+-20) and a scattered pattern from 250+.
More so we make the following observations on the single predictor vs residual plots : 


-  residual vs Temperature : most residuals fall within the 0.2-0.8 temperatures and the plot has a bell shape
-  residual vs Wind speed : we see a funnel pattern with most residuals being in the lower wind speeds.
-  residual vs Humidity :  residuals are scattered with no observable patterns
-  residual vs hour of the day : we observe a bi modal pattern that resemble the rush hours ( opening and closing time )
For residuals vs dummy variables of isWinter and isWorking day :
-  residual vs Winter : we observe both winter and not winter residuals are disturbuted around zero with IQR for not-winter residuals being (400:-400)) while IQR for winter residuals being (-200: 200). both 

#Step 3 
Since a pattern seems to be present in residual vs predictors and residual vs fitted , we need to add new terms or transform predictor variables or response variables

```{R}
# Perform backward selection based on BIC
backwardBICModel <- step(initialFitted, direction = "backward", k = log(nrow(hourData)))

# Display the summary of the selected model
summary(backwardBICModel)

# Plot the residuals vs fitted values for the final model
plot(backwardBICModel, which = 1)  # Residuals vs Fitted

# Plot normal Q-Q plot
plot(backwardBICModel, which = 2)  # Normal Q-Q plot

```